---
title: "week 3 - class 2 - Tidying data - part 2"
author: "Phileas Dazeley Gaist"
date: "10/1/2021"
output: 
  html_document:
    df_print: paged
    keep_md: true
---

```{r}
# # ---------------------------------------------------------------
# ┏━━━┓╋╋┏┓┏┓╋╋╋╋╋╋╋╋╋╋╋╋╋╋┏━┓╋┏┓┏┓╋╋╋╋╋┏━━━┓┏┓┏┓╋╋╋╋╋╋┏┓
# ┃┏━┓┃╋╋┃┃┃┃╋╋╋╋╋╋╋╋╋╋╋╋╋╋┃┏┛┏┛┗┫┃╋╋╋╋╋┃┏━┓┣┛┗┫┃╋╋╋╋╋┏┛┗┓
# ┃┃╋┗╋━━┫┃┃┃┏━━┳━━┳━━┓┏━━┳┛┗┓┗┓┏┫┗━┳━━┓┃┃╋┃┣┓┏┫┃┏━━┳━╋┓┏╋┳━━┓
# ┃┃╋┏┫┏┓┃┃┃┃┃┃━┫┏┓┃┃━┫┃┏┓┣┓┏┛╋┃┃┃┏┓┃┃━┫┃┗━┛┃┃┃┃┃┃┏┓┃┏┓┫┃┣┫┏━┛
# ┃┗━┛┃┗┛┃┗┫┗┫┃━┫┗┛┃┃━┫┃┗┛┃┃┃╋╋┃┗┫┃┃┃┃━┫┃┏━┓┃┃┗┫┗┫┏┓┃┃┃┃┗┫┃┗━┓
# ┗━━━┻━━┻━┻━┻━━┻━┓┣━━┛┗━━┛┗┛╋╋┗━┻┛┗┻━━┛┗┛╋┗┛┗━┻━┻┛┗┻┛┗┻━┻┻━━┛
# ╋╋╋╋╋╋╋╋╋╋╋╋╋╋┏━┛┃
# ╋╋╋╋╋╋╋╋╋╋╋╋╋╋┗━━┛       Applied Data Science I - Week 3, Class 2
# # ---------------------------------------------------------------
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**What makes a data set tidy ?**

- Every column in a data set is a variable
- Every row is an observation
- Every cell is a distinct element (perfect grid)

Often, "untidy" formats are super simple for data entry and collection. 

You'll use pivot longer a lot to prepare data to plot.
You'll use pivot wider a lot when you need to prepare data for analysis, tidy things up. 

```{r}
# Today we are going to learn about just TWO verbs!
#
# "pivot_wider" and "pivot_longer"
# 
# We're also going to reinforce the tidy data principles - namely, a dataset is tidy IF and ONLY IF: 
# 
# Every column is a VARIABLE
# Every row is an OBSERVATION
# Every cell is a SINGLE VALUE 

# ---------------------------------------------------------------
# Verb #1: pivot_longer()
# ---------------------------------------------------------------

library(tidyverse)
data("relig_income")
relig_income

# What makes this dataset untidy?
# ...
# ...
# The column names are are VALUES, not VARIABLES 

# What if I wanted to know how many people identified as agnostic in this dataset? 

sum(relig_income[1,2:11])

# What if I wanted to know how many people identified as agnostic OR atheist but NOT catholic and made at least 75k a year? 
# ...it would be annoying!
# So let's tidy this data up! 

relig_income %>% 
  pivot_longer(!religion, names_to = "income", values_to = "count")

# translation: ignore the religion column, 
# take all values from every column and make a single column called income with every row containing income values for each religious group.
# take all count values from the input data set and put them in a single new column called "count" 

# Let's actually answer the question above 

relig_income %>% 
  pivot_longer(!religion, names_to = 'income', values_to = 'count') %>% 
  filter(religion != 'Catholic' & religion %in% c('Agnostic','Atheist'),
         income %in% c('$75-100k','$100-150k')) %>%
  summarise(count = sum(count))
```

```{r}
# ---------------------------------------------------------------
# Verb #2: pivot_wider
# ---------------------------------------------------------------

data(us_rent_income)
us_rent_income

# hmmmm, this is not tidy! Let's tidy it up. 

us_rent_income %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe))

# now we can answer questions about it!
```

```{r}
# ---------------------------------------------------------------
# General Tidy Data Practice! 
# ---------------------------------------------------------------

data(billboard)
billboard

# Let's clean this up! 

billboard %>% 
  pivot_longer(
    wk1:wk76
  )

billboard %>% 
  pivot_longer(
    # this identifies which rows we want to pivot!
    wk1:wk76, 
    # this tells us the new name of the column
    names_to = "week", 
    # this tells us which values to focus on!
    values_to = "rank"
  )

# That's much better! Let's go even a little farther ...

billboard_cleaned <- billboard %>% 
  pivot_longer(
    wk1:wk76, 
    names_to = "week", 
    values_to = "rank", 
    values_drop_na = TRUE
  ) %>% 
  mutate(
    # this cleans up the "week" column and turns it into an integer - helpful for ordering! 
    week = as.integer(gsub("wk", "", week)),
    # this makes sure that the "date.entered" column becomes a date! 
    date = as.Date(date.entered) + 7 * (week - 1),
  ) %>% 
  # this drops the now unnecessary date column! 
  select(-date.entered)

# Let's answer a couple of questions about the data:

# Which artists had the best average rank in this data set?

billboard_cleaned %>% 
  group_by(artist) %>% 
  summarise(avg_by_artist = mean(rank)) %>% 
  arrange(avg_by_artist)

# Which artist was in the charts for the longest time?

billboard_cleaned %>% 
  group_by(artist) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count)) 

# Which artist has the largest number of songs on the billboard?

billboard_cleaned %>% 
  group_by(artist, track) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count)) %>% 
  ungroup() %>% # Yes, you can ungroup things in the tidyverse.
  group_by(artist) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))

billboard_cleaned %>% 
  group_by(artist) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
  
# How many songs are there by Jay-Z?

billboard_cleaned %>% 
  filter(artist == "Jay-Z") %>% 
  select(track) %>% 
  distinct()
```

**gsub function:** gsub(pattern, thing you want to replace pattern with, vector of strings that you want to replace things in)

```{r}
# ---------------------------------------------------------------
# Let's practice some more! 
# ---------------------------------------------------------------

devtools::install_github("uc-cfss/rcfss")

library(rcfss)


data(grades)

# what's wrong with this dataset? 

# one way to fix it! 

grades %>% 
  pivot_longer(
    Fall:Winter,
    names_to = "Season",
    values_to = "Score"
  )

# ahhh, but we still have an issue - we have multiple rows per observation (see row 1 and row 7). One more step! 

grades %>% 
  pivot_longer(
    Fall:Winter,
    names_to = "Season",
    values_to = "Score"
  ) %>% 
  pivot_wider(
    names_from = Test,
    values_from = Score
  )
```
